---
title: "Coursera"
author: "Charles McKay"
date: "1/23/2017"
output: html_document
---

## Reproducible Research: Week 2 Assignment
### Setup r environment
Load packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r environment}
install.packages("timeDate", repos = "http://cran.us.r-project.org")
library(timeDate)
library(plyr)
library(dplyr)
library(data.table)
```

### Load and preprocess the data
* Read in file
* Add "time" column with converted "interval" column values to time format
* Add "day" column to dipllay week day based on date column values
* Add "wday" logical column TRUE FALSE for is day a weekend on not

```{r process}
step_data <- read.csv("activity.csv")
step_data$time <- substr(as.POSIXct(sprintf("%04.0f", step_data$interval), format='%H%M'), 12, 16)
step_data$time <- as.POSIXct(step_data$time,format="%H:%M")
step_data$day <- weekdays(as.Date(step_data$date))
step_data$subset_key <- paste(step_data$day, step_data$interval, sep="_")
```

## Assigment Questions
### What is mean total number of steps taken per day?

Calculate total steps per day and mean and median steps per day

```{r calculate}
steps_day <- group_by(step_data, date)
steps_sum <- summarize(steps_day, day_sum = sum(steps, na.rm = TRUE))
steps_sum_mean <- mean(steps_sum$day_sum)
steps_sum_median <- median(steps_sum$day_sum)
```

Display histogram of total steps per day

```{r histogram, echo=FALSE}
hist(steps_sum$day_sum, breaks = 10, col = "green", 
     main="Histogram of Total Steps Per Day", xlab = "Total Steps Per Day")
```

Mean and median of the total number of steps taken per day
```{r}
## Mean steps per day
steps_sum_mean

## Median steps per day
steps_sum_median
```

### What is the average daily activity pattern?  
```{r steps by time}
steps_interval <- group_by(step_data, time)
steps_by_time <- summarize(steps_interval, time_mean = mean(steps, na.rm = TRUE))
plot(steps_by_time$time, steps_by_time$time_mean, main="Average Steps Per 5 Minute Interval",
        xlab="Time of Day", ylab="Average Steps", col = "red", type = "l")
```

Maximum average steps steps occur at time interval:
```{r max}
step_time_max <- steps_by_time[which.max(steps_by_time$time_mean),]
format(step_time_max$time, "%H:%M")
```

### Imputing missing values
1. Total number of rows with NAs
```{r NAs}
sum(is.na(step_data$steps))
```
2. New dataset with NAs replaced by imputed values
* Create subset DF for NA rows
* Create subset DF excluding NA rows
* Calculate imputed value as average steps per day/interval combination
* Substitute imputed value for NAs based on day/interval key
* Combine xNA and NA imputed data file to create new DF
```{r add imputed values}
step_data_NA <- step_data[is.na(step_data$steps),]
step_data_xNA <- step_data[!is.na(step_data$steps),]
step_impute <- group_by(step_data, subset_key)
step_impute_mean <- summarize(step_impute, impute_val = mean(steps, na.rm = TRUE))
new_step_data <- merge(step_data_NA[,c("date", "interval" , "time", "day", 
        "subset_key")], step_impute_mean, by= "subset_key")
colnames(new_step_data)[6] <- "steps"
new_step_data <- new_step_data[c(6,2,3,4,5,1)]
new_step_data <- rbind(new_step_data, step_data_xNA)

```

3. Make a histogram of the total number of steps taken each day and
calculate and report the mean and median total number of steps taken per day
```{r calc with imputed data}
new_steps_day <- group_by(new_step_data, date)
new_steps_sum <- summarize(new_steps_day, new_day_sum = sum(steps))
new_steps_sum_mean <- mean(new_steps_sum$new_day_sum)
new_steps_sum_median <- median(new_steps_sum$new_day_sum)
```
* Histogram of total steps taken per day
```{r new histogram}
hist(new_steps_sum$new_day_sum, breaks = 10, col = "blue", 
     main="Histogram of Total Steps Per Day (Imputed)", xlab = "Total Steps Per Day (Imputed)")
```

* Mean and median total number of steps taken per day
```{r mean and median}
## Mean steps per day
new_steps_sum_mean

## Median steps per day
new_steps_sum_median
```

* Differences with the estimates from the first part of the assignment 
After adding imputed values mean and median increased
```{r differenced}
## Mean steps per day difference
new_steps_sum_mean - steps_sum_mean
## Median steps per day difference
new_steps_sum_median - steps_sum_median
```

* Impact of imputing missing data on the estimates of the total daily number of steps

### Are there differences in activity patterns between weekdays and weekends?
* Add weekday/weekend factor
```{r weekday factor}
new_step_data$wday = ifelse(new_step_data$day %in% c("Saturday", "Sunday"), "weekend", "weekday")
new_step_data$wday <- as.factor(new_step_data$wday)
```
* Create panel plot to compare weekday avtivity vs. weekend activity
```{r panel plot}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
## weekday plot
new_step_data_week <- filter(new_step_data, wday == "weekday")
new_steps_interval_week <- group_by(new_step_data_week, time)
new_steps_by_time_week <- summarize(new_steps_interval_week, new_time_mean_week = mean(steps))
plot(new_steps_by_time_week$time, new_steps_by_time_week$new_time_mean_week,
     main="Average Weekday Steps", xlab="Time of Day", ylab="Average Steps", 
     col = "red", type = "l")
## weekend plot
new_step_data_wend <- filter(new_step_data, wday == "weekend")
new_steps_interval_wend <- group_by(new_step_data_wend, time)
new_steps_by_time_wend <- summarize(new_steps_interval_wend, new_time_mean_wend = mean(steps))
plot(new_steps_by_time_wend$time, new_steps_by_time_wend$new_time_mean_wend,
     main="Average Weekend Steps", xlab="Time of Day", ylab="Average Steps", 
     col = "blue", type = "l")
```

* Weekend activity ramps up more slowly in the morning and is higher post 10:00 


